The Ky Fan Norms and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-12-01 article_final/main.tex
many changes for notational consistency and clarity

% Norms and operators
\newcommand{\norm}[1]{\lVert #1\rVert}
\DeclarePairedDelimiter{\normf}{\|}{\|_\mathrm{F}}
\DeclarePairedDelimiter{\norms}{\|}{\|_{\mathrm{2}}}
\DeclarePairedDelimiter{\normc}{\|}{\|_{\mathrm{C}}}
\DeclarePairedDelimiter{\normn}{\|}{\|_{*}}
\newcommand{\normkfk}[1]{\|#1\|_{\text{KF-}k}}
\DeclarePairedDelimiter{\normfstar}{\|}{\|_\mathrm{F*}}
\DeclarePairedDelimiter{\normftwo}{\|}{\|_\mathrm{F2}}
\DeclarePairedDelimiter{\abs}{\lvert}{\rvert}

@@

\begin{lemma}\label{lemma:dual_to_conv_comb}
	Let $\norm{\cdot}_{(1)},\dots,\norm{\cdot}_{(n)}$ be norms on a finite-dimensional Euclidean space, and let $\alpha_1,\dots,\alpha_n$ be non-negative reals. Define
	\[
		\norm{\cdot} := \sum_{i = 1}^n \alpha_i \norm{\cdot}_{(i)}.
	\]
	Then the dual unit ball of $\norm{\cdot}$ satisfies
	\[
		\cB_{\norm{\cdot}^\dagger}
		= \sum_{i = 1}^n \alpha_i \cB_{\norm{\cdot}_{(i)}^\dagger}
	\]
	where $\sum$ denotes the Minkowski sum and $\cB_{\norm{\cdot}_{(i)}^\dagger}$ is the unit ball of the dual norm $\norm{\cdot}_{(i)}^\dagger$.
\end{lemma}

@@

\begin{table}[ht]
\floatconts
  {tbl:matrix_methods}
  {\caption{Comparison of methods for computing rank-$k$ updates on dense random matrices (CPU, double precision). Lower is better in all columns.}}
  {%
    \begin{tabular}{|c|c|c|c|c|c|c|c|}
      \arrayrulecolor{black}\hline
      Matrix sizes & $k$  & Method           & Time (s) & Matvecs & Iterations & $err_1$         & $err_2$       \\
      \arrayrulecolor{black}\hline
      $500 \times 500$      & 5  & Power Iterations           & 0.062   & 2005    & 200        & 9.2$\cdot 10^{-3}$ & 9.1$\cdot10^{-3}$\\
      $500 \times 500$      & 5  & RSVD             & 0.017   & 1170    & 38         & 9.8$\cdot 10^{-3}$ & 9.6$\cdot10^{-3}$\\
      $500 \times 500$      & 5  & TRLan            & 0.018   & 131     & 65         & 9.6$\cdot 10^{-5}$ & 9.4$\cdot10^{-5}$\\
      \hdashline
      $500 \times 500$      & 50 & Power Iterations           & 0.44    & 43750   & 437      & 9.9$\cdot 10^{-3}$ & 9.0$\cdot10^{-3}$\\
      $500 \times 500$      & 50 & RSVD             & 0.61    & 6120    & 50      & 9.9$\cdot 10^{-3}$ & 9.1$\cdot10^{-3}$\\
      $500 \times 500$      & 50 & TRLan            & 0.16    & 462     & 231      & 3.3$\cdot 10^{-7}$ & 3.0$\cdot10^{-7}$\\
      \hdashline 
      $5000 \times 5000$    & 5  & Power Iterations           & 9.6     & 9065    & 906      & 8.6$\cdot 10^{-3}$ & 8.6$\cdot10^{-3}$\\
      $5000 \times 5000$    & 5  & RSVD             & 2.1     & 5640    & 187      & 9.7$\cdot 10^{-3}$ & 9.7$\cdot10^{-3}$\\
      $5000 \times 5000$    & 5  & TRLan            & 0.70    & 205     & 102      & 7.7$\cdot 10^{-3}$ & 7.7$\cdot10^{-3}$\\
      \arrayrulecolor{black}\hline
    \end{tabular}%
  }
\end{table}

\begin{table}[ht]
\floatconts
  {tbl:newton_schulz_reference}
  {\caption{Newton-Schulz iterations on the same matrices (for reference).}}
  {%
    \begin{tabular}{|c|c|c|c|c|}
      \arrayrulecolor{black}\hline
      Matrix size    & Time (s) & Matvecs & Iterations & $err_1$ \\
      \arrayrulecolor{black}\hline
      $500 \times 500$     & 0.041    & 27 000  & 27         & 4.8e-3  \\
      $5000 \times 5000$   & 26.4     & 290 000 & 29         & 6.5e-3  \\
      \arrayrulecolor{black}\hline
    \end{tabular}%
  }
\end{table}

@@

We begin by evaluating F-Fanions on the following convex $L$-smooth problem:
\begin{equation}\label{eq:lls}
  F(\mX) = \frac{1}{2} \<(\mX-\mS), \mM (\mX - \mS) \mN> \to \min_{\mX \in \Rmn}\,
\end{equation}
where $\mX \in \Rmn$, $m=500$ and $n=500$ represent typical dimensions of a neural network weight matrix, $\mS \in \Rmn$, and $\mM \in \mathbb{S}^m_{+}$ and $\mN \in \mathbb{S}^n_{+}$ are positive-semidefinite matrices. The spectra of $\mM$ and $\mN$ are uniformly distributed on the interval $(0, 1)$. We set $\mS=0$ and initialize $\mX^0$ with entries drawn independently from $\cN(0, 0.01)$.

@@

We perform a grid search over the following hyperparameter ranges:
\begin{itemize}
  \item \texttt{momentum} $\in \{0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 0.95\}$ for all algorithms.
  \item \texttt{lr} $\in [0.005, 0.020]$ with step $0.001$ for Muon, F-Muon, and S-Muon.
  \item \texttt{lr} $\in [5\mathrm{e}{-5},\, 2\mathrm{e}{-4}]$ with step $1\mathrm{e}{-5}$.
  \item \texttt{lr} $\in [0.01, 0.10]$ with step $0.001$ for NSGD.
\end{itemize}
The tuned parameters and the number of iterations required to converge to $0.001$ are presented in \tableref{tbl:lls_lrs}. For intermediate-rank Fanions, F-Fanions, and S-Fanions, the hyperparameters \texttt{lr}, \texttt{momentum}, and \texttt{sign\_lr\_coeff} are set equal to those of Muon, F-Muon, and S-Muon respectively, as their loss decreases too slowly to reach $0.001$ within a reasonable tuning budget.

\begin{table}[htbp]
\floatconts
  {tbl:lls_lrs}
  {\caption{Tuning learning rates and momentum coefficients}}
  {%
    \begin{tabular}{|c|c|c|c|}
      \arrayrulecolor{black}\hline
      Algorithm & \texttt{lr}  & \texttt{momentum}  & Iterations to 0.001 loss \\
      \arrayrulecolor{black}\hline
      Muon & 0.007 & 0.5 & 1060 \\
      \hdashline
      NSGD & 0.08 & 0.95 & 1020 \\
      F-Muon & 0.015 & 0.7 & 910 \\
      \hdashline
      SignSGD & 0.016 $\times$ 0.01 & 0.95 & 2650 \\
      S-Muon & 0.011 & 0.9 & 890 \\
      \arrayrulecolor{black}\hline
    \end{tabular}%
  }
\end{table}

@@

\begin{figure}[htbp]
\floatconts
  {fig:lls}
  {\caption{Linear least squares problem for a 500x500 matrix}}
  {%
    \subfigure[The loss]{\label{fig:lls_loss}%
      \includegraphics[width=0.8\linewidth]{figs/simple_lls/loss_long_run_16_increase.pdf}}%
    \vspace{1em}
    \subfigure[The Frobenius norm of the gradient]{\label{fig:lls_fro_grad_norm}%
      \includegraphics[width=0.8\linewidth]{figs/simple_lls/gradient_frobenius_norm_vs_iteration_long_run.pdf}}
  }
\end{figure}

@@

We evaluate the algorithms on the CIFAR-10 airbench \citep{cifar2023airbench}. To assess the impact of the mixing parameter $\alpha$, we first run F-Muon for different values of $\alpha$ using hyperparameters tuned for vanilla Muon by Keller Jordan: \texttt{lr=0.24(1 - step/total\_steps), momentum=0.65, nesterov=True} with weight normalization. We perform 10 repetitions for each $\alpha$ value and record the accuracy after 8 epochs of training (\figureref{fig:muon_alphas}).

Next, we tune F-Muon specifically with $\alpha = 0.5$, obtaining optimal hyperparameters \texttt{lr=0.4(1 - step/\allowbreak{}total\_steps)}, \texttt{momentum=0.6}, \texttt{nesterov = True}. Tuned F-Muon achieves $94.02 \pm 0.13\%$ validation accuracy after 8 epochs (averaged over 200 runs), matching Muon's accuracy and variance. We again measure the validation accuracy as a function of $\alpha$ (\figureref{fig:fmuon_alphas}) and observe that even at $\alpha=0.1$, the accuracy substantially exceeds that of vanilla NSGD.

For S-Muon with $\alpha=0.5$, we tune all hyperparameters to find the optimal configuration \texttt{lr=0.42(1 - step/\allowbreak{}total\_steps)}, \texttt{momentum=0.63}, \texttt{nesterov = True}, \texttt{sign\_lr\_coeff = 0.003}, achieving $94.03 \pm 0.13\%$ validation accuracy, which slightly exceeds vanilla Muon's performance.

\begin{figure}[htbp]
\floatconts
  {fig:diff_alphas}
  {\caption{Mean validation accuracies for F-Muon with different $\alpha$}}
  {%
    \subfigure[With parameters tuned for Muon]{\label{fig:muon_alphas}%
      \includegraphics[width=0.48\linewidth]{figs/muon_tuned_diff_alpha.pdf}}%
    \hfill
    \subfigure[With parameters tuned for F-Muon]{\label{fig:fmuon_alphas}%
      \includegraphics[width=0.48\linewidth]{figs/fmuon_tuned_diff_alpha.pdf}}
  }
\end{figure}

The Muon-level performance is remarkable given that F-Muon and S-Muon operate with substantially different constraint geometries. \figureref{fig:cifar_ball} visualizes F-Muon by plotting the LMO balls of Muon and F-Muon (scaled by actual learning rates) in the 2D space of singular values. S-Muon cannot be visualized in this manner because the $l_\infty$ norm is not a function of singular values alone, and visualizing $\norms{\cdot}$ for $\Rmn$ with $m, n > 2$ requires at least 4D space. Nevertheless, the LMO ball of S-Muon clearly differs substantially from Muon's, as the effective SignSGD learning rate $\texttt{lr}(1-\alpha)\texttt{sign\_lr\_coeff} = 6.3 \times 10^{-4}$ is comparable to typical SignSGD learning rates in deep learning.

Notably, even the pathological case $\alpha > 1$, which corresponds to an LMO constraint region that is not a norm ball, achieves accuracy nearly identical to vanilla Muon.

These observations raise fundamental questions about the sensitivity of LMO-based algorithms to the shape of the constraint region.

For gradient norm profiling and results for training with Fanions and F-Fanions, please see \sectionref{sec:cifar_underhood}.

@@

@@

We evaluate F-Muon and S-Muon with $\alpha=0.5$ against Muon, SignSGD, and NSGD on the NanoGPT speedrun benchmark \citep{modded_nanogpt_2024}. The optimal hyperparameters are shown in the legend of \figureref{fig:nano_gpt_loss}, with \texttt{sign\_sgd\_coeff} values of 3e-4/0.07 for S-Muon. After 1750 training steps, F-Muon achieves a cross-entropy loss of 3.281, S-Muon achieves 3.287, and Muon achieves 3.279, falling below the target threshold of 3.280. As illustrated in \figureref{fig:nano_gpt_loss}, these differences are negligible. Notably, F-Muon represents a convex combination of Muon and NSGD, despite NSGD performing poorly in isolation. The same observation holds for S-Muon.

\begin{figure}[htbp]
\floatconts
  {fig:nano_gpt_loss}
  {\caption{The validation loss for NanoGPT}}
  {\includegraphics[width=0.7\linewidth]{figs/nanogpt/nano/val_loss_vs_step.pdf}}
\end{figure}

@@

\subsection{GPT-2 Medium Speedrun}
We scale from NanoGPT to GPT-2 Medium. After testing for 5960 steps we get 2.9215 loss for F-Muon and 2.9235 for S-Muon, while for Muon it is 2.9198 (the speedrun threshold is 2.92).

\begin{figure}[htbp]
  \floatconts
    {fig:med_gpt_loss}
    {\caption{The validation loss for GPT-2 Medium}}
    {\includegraphics[width=0.7\linewidth]{figs/nanogpt/medium/val_loss_vs_step.pdf}}
  \end{figure}

@@

We fine-tuned the NanoGPT framework by Karpathy \citep{Karpathy2022} using the standard \emph{NanoGPT-medium} configuration, which corresponds to GPT-2 Medium. This model consists of 24 transformer layers, a hidden dimension of 1024, 16 attention heads, and approximately 345 million parameters. We selected the \emph{Tiny Stories} corpus \citep{eldan2023tinystoriessmalllanguagemodels} as the training dataset for its high entropy and structural diversity, which amplify the differences in optimization dynamics. All training runs were initialized with the same random seed, weight initialization, and learning rate schedule to ensure that performance differences arose solely from the choice of optimizer.

For all one-dimensional layers, we used AdamW with a learning rate of $1\!\times\!10^{-3}$. Since momentum exhibited negligible influence on training dynamics, it was held constant across all experiments. \figureref{fig:nanogptftexps} presents a comparative analysis of optimization performance.

\begin{figure}[htbp]
\floatconts
  {fig:nanogptftexps}
  {\caption{Comparison of validation loss for Muon, F-Muon, and S-Muon across a range of learning rates. Stars denote the best learning rate identified for each optimizer.}}
  {\includegraphics[width=0.8\linewidth]{figs/nanogpt/optimizers_comparison.pdf}}
\end{figure}

@@

\begin{table}[htbp]
\floatconts
  {tbl:mat_vs_vec_lmo}
  {\caption{LMO optimizers in Schatten $S_p$ norms and in $l_p$ norms. $g$ is the gradient. When it is a matrix, $g = \mU \mSigma \mV^\top$}}
  {%
    \begingroup
    \def\arraystretch{1.2}
    \resizebox{\linewidth}{!}{%
      \begin{tabular}{|l|c|c|c|}
        \arrayrulecolor{black}\hline
        Algorithm                             & LMO constraint set $\mathcal{D}$ & LMO                                                         & Reference                            \\
        \arrayrulecolor{black}\hline
        \arrayrulecolor{black}\hline
        Normalized SGD                     & $l_2$-ball, $S_2$-ball          & $-\eta \tfrac{g}{\norm{g}_2} = -\eta \tfrac{g}{\normf{g}}$ & \citep{hazan2015beyond}              \\
        Momentum Normalized SGD            & Ball in $l_2$, or Ball in $S_2$ & $-\eta \tfrac{g}{\norm{g}_2} = -\eta\tfrac{g}{\normf{g}}$  & \citep{cutkosky2020momentum}         \\
        \arrayrulecolor{black}\hline
        SignSGD                            & Ball in Max-norm $l_\infty$     & $-\eta \sign(g)$                                            & \citep[Thm. 1]{bernstein2018signsgd} \\
        Signum                             & Ball in Max-norm $l_\infty$     & $-\eta \sign(g)$                                            & \citep[Thm. 3]{bernstein2018signsgd} \\
        \hdashline
        Muon                               & Ball in Spectral $S_\infty$     & $-\eta \mU\mV^\top$                                             & \citep{jordan2024muon}               \\
        \arrayrulecolor{black}\hline
        Gauss-Southwell Coordinate Descent & Ball in $l_1$                   & $-\eta \sum_{i \in \argmax|g_i^t|} \sign(g_i^t)e_i$                       & \citep[p. 19]{shi2016primer}         \\
        \hdashline
        Neon                               & Ball in Nuclear $S_1$           & $-\eta \vu_1 \vv_1^\top$                                        & This work                            \\
        \arrayrulecolor{black}\hline
      \end{tabular}%
    }%
    \endgroup
  }
\end{table}

@@

Based on the \lemmaref{lemma:dual_to_conv_comb}, we immediately find $\normfstar{\cdot}^\dagger$, which is related to F-Muon update. Indeed, after setting $\beta=1-\alpha$ and remembering that for smooth and bounded cases we can use $\min$ instead of $\inf$, we get
\begin{equation}\label{eq:normfstardual}
  \normfstar{\mY}^\dagger = \min_{\mZ} \min_t\{t, s.t. \norms{\mZ}\leq \alpha t, \normf{\mY-\mZ}\leq (1-\alpha) t\}
\end{equation}

@@

Similarly, we find $\normftwo{\cdot}^\dagger$, which is related to F-Neon update:
\begin{equation}\label{eq:normftwodual}
  \normftwo{\mY}^\dagger = \min_{\mZ} \min_t\{t, s.t. \normn{\mZ}\leq \alpha t, \normf{\mY-\mZ}\leq (1-\alpha) t\}
\end{equation}

@@

Unfortunately, $\normfstar{\cdot}^\dagger$ and $\normftwo{\cdot}^\dagger$ do not have closed-form expressions and cannot be found as easy as their duals.

@@

In \figureref{fig:fduals} we plot these norms. On x-axis and y-axis, there are singular values $\sigma_1$, $\sigma_2$ respectively of a matrix from $\R^{m\times n}$ with $\min\{m,n\}=2$.

>>>>

% Norms and operators
\newcommand{\norm}[1]{\lVert #1\rVert}
\DeclarePairedDelimiter{\normf}{\|}{\|_\mathrm{F}}
\DeclarePairedDelimiter{\norms}{\|}{\|_{\mathrm{2}}}
\DeclarePairedDelimiter{\normc}{\|}{\|_{\mathrm{C}}}
\DeclarePairedDelimiter{\normn}{\|}{\|_{*}}
\DeclarePairedDelimiter{\normkfk}{\|}{\|_{\text{KF-}k}}
\DeclarePairedDelimiter{\normfstar}{\|}{\|_\mathrm{F*}}
\DeclarePairedDelimiter{\normftwo}{\|}{\|_\mathrm{F2}}
\DeclarePairedDelimiter{\abs}{\lvert}{\rvert}

@@

\begin{lemma}\label{lemma:dual_to_conv_comb}
	Let $\norm{\cdot}_{(1)},\dots,\norm{\cdot}_{(n)}$ be norms on a finite-dimensional Euclidean space, and let $\alpha_1,\dots,\alpha_n$ be non-negative reals. Define
	\[
		\norm{\cdot} := \sum_{i = 1}^n \alpha_i \norm{\cdot}_{(i)}\,.
	\]
	Then the dual unit ball of $\norm{\cdot}$ satisfies
	\[
		\cB_{\norm{\cdot}^\dagger}
		= \sum_{i = 1}^n \alpha_i \cB_{\norm{\cdot}_{(i)}^\dagger}\,,
	\]
	where $\sum$ denotes the Minkowski sum and $\cB_{\norm{\cdot}_{(i)}^\dagger}$ is the unit ball of the dual norm $\norm{\cdot}_{(i)}^\dagger$.
\end{lemma}

@@

\begin{table}[ht]
\floatconts
  {tbl:matrix_methods}
  {\caption{Comparison of methods for computing rank-$k$ updates on dense random matrices (CPU, double precision). Lower is better in all columns.}}
  {%
    \begin{tabular}{|c|c|c|c|c|c|c|c|}
      \arrayrulecolor{black}\hline
      Matrix sizes & $k$  & Method           & Time (s) & Matvecs & Iterations & $err_1$         & $err_2$       \\
      \arrayrulecolor{black}\hline
      $500 \times 500$      & 5  & Power Iterations           & 0.062   & 2005    & 200        & $9.2 \cdot 10^{-3}$ & $9.1 \cdot 10^{-3}$\\
      $500 \times 500$      & 5  & RSVD             & 0.017   & 1170    & 38         & $9.8 \cdot 10^{-3}$ & $9.6 \cdot 10^{-3}$\\
      $500 \times 500$      & 5  & TRLan            & 0.018   & 131     & 65         & $9.6 \cdot 10^{-5}$ & $9.4 \cdot 10^{-5}$\\
      \hdashline
      $500 \times 500$      & 50 & Power Iterations           & 0.44    & 43750   & 437      & $9.9 \cdot 10^{-3}$ & $9.0 \cdot 10^{-3}$\\
      $500 \times 500$      & 50 & RSVD             & 0.61    & 6120    & 50      & $9.9 \cdot 10^{-3}$ & $9.1 \cdot 10^{-3}$\\
      $500 \times 500$      & 50 & TRLan            & 0.16    & 462     & 231      & $3.3 \cdot 10^{-7}$ & $3.0 \cdot 10^{-7}$\\
      \hdashline
      $5000 \times 5000$    & 5  & Power Iterations           & 9.6     & 9065    & 906      & $8.6 \cdot 10^{-3}$ & $8.6 \cdot 10^{-3}$\\
      $5000 \times 5000$    & 5  & RSVD             & 2.1     & 5640    & 187      & $9.7 \cdot 10^{-3}$ & $9.7 \cdot 10^{-3}$\\
      $5000 \times 5000$    & 5  & TRLan            & 0.70    & 205     & 102      & $7.7 \cdot 10^{-3}$ & $7.7 \cdot 10^{-3}$\\
      \arrayrulecolor{black}\hline
    \end{tabular}%
  }
\end{table}

\begin{table}[ht]
\floatconts
  {tbl:newton_schulz_reference}
  {\caption{Newton-Schulz iterations performance on random dense matrices (for reference).}}
  {%
    \begin{tabular}{|c|c|c|c|c|}
      \arrayrulecolor{black}\hline
      Matrix size    & Time (s) & Matvecs & Iterations & $err_1$ \\
      \arrayrulecolor{black}\hline
      $500 \times 500$     & 0.041    & 27 000  & 27         & $4.8 \cdot 10^{-3}$  \\
      $5000 \times 5000$   & 26.4     & 290 000 & 29         & $6.5 \cdot 10^{-3}$  \\
      \arrayrulecolor{black}\hline
    \end{tabular}%
  }
\end{table}

@@

We begin by evaluating F-Fanions on the following convex $L$-smooth problem:
\begin{equation}\label{eq:lls}
  F(\mX) = \frac{1}{2} \<(\mX-\mS), \mM (\mX - \mS) \mN> \to \min_{\mX \in \Rmn}\,,
\end{equation}
where $\mX \in \Rmn$, $m=500$ and $n=500$ represent typical dimensions of a neural network weight matrix, $\mS \in \Rmn$, and $\mM \in \mathbb{S}^m_{+}$ and $\mN \in \mathbb{S}^n_{+}$ are positive-semidefinite matrices. The spectra of $\mM$ and $\mN$ are uniformly distributed on the interval $(0, 1)$. We set $\mS=0$ and initialize $\mX^0$ with entries drawn independently from $\cN(0, 0.01)$.

@@

We perform a grid search over the following hyperparameter ranges:
\begin{itemize}
  \item \texttt{momentum} $\in \{0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 0.95\}$ for all algorithms.
  \item \texttt{lr} $\in [0.005, 0.020]$ with step $0.001$ for Muon, F-Muon, and S-Muon.
  \item \texttt{lr} $\in [5 \cdot 10^{-5},\, 2 \cdot 10^{-4}]$ with step $1 \cdot 10^{-5}$.
  \item \texttt{lr} $\in [0.01, 0.10]$ with step $0.001$ for NSGD.
\end{itemize}
The tuned parameters and the number of iterations required to converge to $0.001$ are presented in \tableref{tbl:lls_lrs}. For intermediate-rank Fanions, F-Fanions, and S-Fanions, the hyperparameters \texttt{lr}, \texttt{momentum}, and \texttt{sign\_lr\_coeff} are set equal to those of Muon, F-Muon, and S-Muon respectively, as their loss decreases too slowly to reach $0.001$ within a reasonable tuning budget.

\begin{table}[htbp]
\floatconts
  {tbl:lls_lrs}
  {\caption{Tuned learning rates and momentum coefficients in the linear least squares experiment.}}
  {%
    \begin{tabular}{|c|c|c|c|}
      \arrayrulecolor{black}\hline
      Algorithm & \texttt{lr}  & \texttt{momentum}  & Iterations to 0.001 loss \\
      \arrayrulecolor{black}\hline
      Muon & 0.007 & 0.5 & 1060 \\
      \hdashline
      NSGD & 0.08 & 0.95 & 1020 \\
      F-Muon & 0.015 & 0.7 & 910 \\
      \hdashline
      SignSGD & 0.016 $\times$ 0.01 & 0.95 & 2650 \\
      S-Muon & 0.011 & 0.9 & 890 \\
      \arrayrulecolor{black}\hline
    \end{tabular}%
  }
\end{table}

@@

\begin{figure}[htbp]
\floatconts
  {fig:lls}
  {\caption{Linear least squares problem for a 500x500 matrix.}}
  {%
    \subfigure[The loss]{\label{fig:lls_loss}%
      \includegraphics[width=0.8\linewidth]{figs/simple_lls/loss_long_run_16_increase.pdf}}%
    \vspace{1em}
    \subfigure[The Frobenius norm of the gradient]{\label{fig:lls_fro_grad_norm}%
      \includegraphics[width=0.8\linewidth]{figs/simple_lls/gradient_frobenius_norm_vs_iteration_long_run.pdf}}
  }
\end{figure}

@@

We evaluate the algorithms on the CIFAR-10 airbench \citep{cifar2023airbench}. To assess the impact of the mixing parameter $\alpha$, we first run F-Muon for different values of $\alpha$ using hyperparameters tuned for vanilla Muon by Keller Jordan: \texttt{lr=0.24(1 - step/total\_steps)}, \texttt{momentum=0.65}, \texttt{nesterov=True} with weight normalization. We perform 10 repetitions for each $\alpha$ value and record the accuracy after 8 epochs of training (\figureref{fig:muon_alphas}).

Next, we tune F-Muon specifically with $\alpha = 0.5$, obtaining optimal hyperparameters \texttt{lr=0.4(1 - step/total\_steps)}, \texttt{momentum=0.6}, \texttt{nesterov = True}. Tuned F-Muon achieves $94.02 \pm 0.13\%$ validation accuracy after 8 epochs (averaged over 200 runs), matching Muon's accuracy and variance. We again measure the validation accuracy as a function of $\alpha$ (\figureref{fig:fmuon_alphas}) and observe that even at $\alpha=0.1$, the accuracy substantially exceeds that of vanilla NSGD.

For S-Muon with $\alpha=0.5$, we tune all hyperparameters to find the optimal configuration \texttt{lr=0.42(1 - step/\allowbreak{}total\_steps)}, \texttt{momentum=0.63}, \texttt{nesterov = True}, \texttt{sign\_lr\_coeff = 0.003}, achieving $94.03 \pm 0.13\%$ validation accuracy (\figureref{fig:diff_alphas}), which slightly exceeds vanilla Muon's performance.

\begin{figure}[htbp]
\floatconts
  {fig:diff_alphas}
  {\caption{Mean validation accuracies for F-Muon with different $\alpha$.}}
  {%
    \subfigure[With parameters tuned for Muon]{\label{fig:muon_alphas}%
      \includegraphics[width=0.48\linewidth]{figs/muon_tuned_diff_alpha.pdf}}%
    \hfill
    \subfigure[With parameters tuned for F-Muon]{\label{fig:fmuon_alphas}%
      \includegraphics[width=0.48\linewidth]{figs/fmuon_tuned_diff_alpha.pdf}}
  }
\end{figure}

The Muon-level performance is remarkable given that F-Muon and S-Muon operate with substantially different constraint geometries. \figureref{fig:cifar_ball} visualizes F-Muon by plotting the LMO balls of Muon and F-Muon (scaled by actual learning rates) in the 2D space of singular values (with all other singular values set to zero). S-Muon cannot be visualized in this manner because the Chebyshev norm is not a function of singular values alone, and visualizing $\norms{\cdot}$ for $\Rmn$ with $m, n > 2$ requires at least 4D space. Nevertheless, the LMO ball of S-Muon clearly differs substantially from Muon's, as the effective SignSGD learning rate $\texttt{lr} = (1-\alpha)\texttt{sign\_lr\_coeff} = 6.3 \cdot 10^{-4}$ is comparable to typical SignSGD learning rates in deep learning.

@@

Notably, even the pathological case $\alpha > 1$, which corresponds to an LMO constraint region that is not a norm ball, achieves accuracy nearly identical to vanilla Muon. These observations raise fundamental questions about the sensitivity of LMO-based algorithms to the shape of the constraint region.

Additional results, including gradient norm profiling and performance comparisons with Fanions and F-Fanions, are provided in \sectionref{sec:cifar_underhood}.

@@

We evaluate F-Muon and S-Muon with $\alpha=0.5$ against Muon, SignSGD, and NSGD on the NanoGPT speedrun benchmark \citep{modded_nanogpt_2024}. The optimal hyperparameters are shown in the legend of \figureref{fig:nano_gpt_loss}, with \texttt{sign\_sgd\_coeff} values of $3 \cdot 10^{-4}$ and $0.07$ for S-Muon. After 1750 training steps, F-Muon achieves a cross-entropy loss of 3.281, S-Muon achieves 3.287, and Muon achieves 3.279, falling below the target threshold of 3.280. As illustrated in \figureref{fig:nano_gpt_loss}, these differences are negligible. Notably, F-Muon represents a convex combination of Muon and NSGD, despite NSGD performing poorly in isolation. The same observation holds for S-Muon.

\begin{figure}[htbp]
\floatconts
  {fig:nano_gpt_loss}
  {\caption{The validation loss for NanoGPT.}}
  {\includegraphics[width=0.7\linewidth]{figs/nanogpt/nano/val_loss_vs_step.pdf}}
\end{figure}

@@

\subsection{GPT-2 Medium Speedrun}

We scale from NanoGPT to GPT-2 Medium (24 transformer layers, 1024 hidden dimension, 16 attention heads, approximately 345 million parameters), evaluating the same algorithms with $\alpha=0.5$ on the FineWeb dataset. After 5960 training steps, Muon achieves a validation loss of 2.9198, successfully reaching the speedrun threshold of 2.92. F-Muon achieves 2.9215 and S-Muon achieves 2.9235, narrowly missing the threshold. As illustrated in \figureref{fig:med_gpt_loss}, the performance gap remains remarkably small: F-Muon trails Muon by only 0.0017, while S-Muon trails by 0.0037. These minimal differences demonstrate that alternative norm constraints maintain competitive performance even at this significantly larger model scale.

\begin{figure}[htbp]
  \floatconts
    {fig:med_gpt_loss}
    {\caption{The validation loss for GPT-2 Medium.}}
    {\includegraphics[width=0.7\linewidth]{figs/nanogpt/medium/val_loss_vs_step.pdf}}
  \end{figure}

@@

We fine-tuned the NanoGPT framework by Karpathy \citep{Karpathy2022} using the standard \emph{NanoGPT-medium} configuration, which again corresponds to GPT-2 Medium. We selected the \emph{Tiny Stories} corpus \citep{eldan2023tinystoriessmalllanguagemodels} as the training dataset for its high entropy and structural diversity, which amplify the differences in optimization dynamics. All training runs were initialized with the same random seed, weight initialization, and learning rate schedule to ensure that performance differences arose solely from the choice of optimizer.

For all one-dimensional layers, we used AdamW with a learning rate of $1 \cdot 10^{-3}$. Since momentum exhibited negligible influence on training dynamics, it was held constant across all experiments. \figureref{fig:nanogptftexps} presents a comparative analysis of optimization performance.

\begin{figure}[htbp]
\floatconts
  {fig:nanogptftexps}
  {\caption{Comparison of validation loss for Muon, F-Muon, and S-Muon across a range of learning rates. Stars denote the best learning rate identified for each optimizer.}}
  {\includegraphics[width=0.7\linewidth]{figs/nanogpt/optimizers_comparison.pdf}}
\end{figure}

@@

\begin{table}[htbp]
\floatconts
  {tbl:mat_vs_vec_lmo}
  {\caption{LMO optimizers in Schatten $S_p$ norms and in $l_p$ norms. $g$ is the gradient. When it is a matrix, $g = \mU \mSigma \mV^\top$.}}
  {%
    \begingroup
    \def\arraystretch{1.2}
    \resizebox{\linewidth}{!}{%
      \begin{tabular}{|l|c|c|c|}
        \arrayrulecolor{black}\hline
        Algorithm                             & LMO constraint set $\mathcal{D}$ & LMO                                                         & Reference                            \\
        \arrayrulecolor{black}\hline
        \arrayrulecolor{black}\hline
        Normalized SGD                     & $l_2$-ball, $S_2$-ball          & $-\eta \tfrac{g}{\norm{g}_2} = -\eta \tfrac{g}{\normf{g}}$ & \citep{hazan2015beyond}              \\
        Momentum Normalized SGD            & Ball in $l_2$, or Ball in $S_2$ & $-\eta \tfrac{g}{\norm{g}_2} = -\eta\tfrac{g}{\normf{g}}$  & \citep{cutkosky2020momentum}         \\
        \arrayrulecolor{black}\hline
        SignSGD                            & Ball in Max-norm $l_\infty$     & $-\eta \sign(g)$                                            & \citep[Thm. 1]{bernstein2018signsgd} \\
        Signum                             & Ball in Max-norm $l_\infty$     & $-\eta \sign(g)$                                            & \citep[Thm. 3]{bernstein2018signsgd} \\
        \hdashline
        Muon                               & Ball in Spectral $S_\infty$     & $-\eta \mU\mV^\top$                                             & \citep{jordan2024muon}               \\
        \arrayrulecolor{black}\hline
        Gauss-Southwell Coordinate Descent & Ball in $l_1$                   & $-\eta \sum_{i \in \argmax|g_i^t|} \sign(g_i^t)e_i$                       & \citep[p. 19]{shi2016primer}         \\
        \hdashline
        Neon                               & Ball in Nuclear $S_1$           & $-\eta \vu_1 \vv_1^\top$                                        & This work                            \\
        \arrayrulecolor{black}\hline
      \end{tabular}%
    }%
    \endgroup
  }
\end{table}

@@

Based on \lemmaref{lemma:dual_to_conv_comb}, we immediately find $\normfstar{\cdot}^\dagger$, which is related to the F-Muon update. Indeed, after setting $\beta=1-\alpha$ and remembering that for smooth and bounded cases we can use $\min$ instead of $\inf$, we obtain
\begin{equation}\label{eq:normfstardual}
  \normfstar{\mY}^\dagger = \min_{\mZ} \min_t\{t, s.t. \norms{\mZ}\leq \alpha t, \normf{\mY-\mZ}\leq (1-\alpha) t\}\,.
\end{equation}

@@

Similarly, we find $\normftwo{\cdot}^\dagger$, which is related to the F-Neon update:
\begin{equation}\label{eq:normftwodual}
  \normftwo{\mY}^\dagger = \min_{\mZ} \min_t\{t, s.t. \normn{\mZ}\leq \alpha t, \normf{\mY-\mZ}\leq (1-\alpha) t\}\,.
\end{equation}

@@

Unfortunately, $\normfstar{\cdot}^\dagger$ and $\normftwo{\cdot}^\dagger$ do not have simple closed-form expressions and cannot be computed as easily as their duals.

@@

In \figureref{fig:fduals}, we plot these norms. The x-axis and y-axis represent the singular values $\sigma_1$ and $\sigma_2$, respectively, of a matrix from $\R^{m\times n}$ with $\min\{m,n\}=2$.

\begin{figure}[htbp]
\floatconts
  {fig:fduals}
  {\caption{Balls in the duals to F* and F2 norms for different $\alpha$.}}
  {%
    \subfigure[lmo balls for F-Muon for different $\alpha$]{\label{fig:fstardual}%
      \includegraphics[width=0.48\linewidth]{figs/norms/fstardualball.pdf}}%
    \hfill
    \subfigure[lmo ball for F-Neon for different $\alpha$]{\label{fig:ftwodual}%
      \includegraphics[width=0.48\linewidth]{figs/norms/ftwodualball.pdf}}
  }
\end{figure}
